---
  title: "ICT513_Project_AGAMPODI"
author: "Supuni Agampodi"
date: "2023-11-03"
output: pdf_document
---
  
  ```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


```{r }
## clear the working directory
rm(list = ls())
library(devtools)
devtools::install_github("vqv/ggbiplot")
# Load the "dplyr" package.
library(dplyr)
# Load the "purrr" package.
library(purrr)
# Load the "ggplot2" package.
library(ggplot2)
# Load the "bootstrap" package.
library(bootstrap)
# Load the "leaps" package.
library(leaps)
library(plyr)
library(scales)
library(gridExtra)
```

# Introduction


## Read Data
```{r}
# Set the working directory to the specified location
setwd("D:/Supuni/Murdoch/Y2S2/ICT513-DA/Project")

# Read the dataset 'SeoulBikeData.csv' into a variable called 'bike'
bike <- read.csv("SeoulBikeData.csv", check.names = FALSE)

# Rename columns for clarity
bike <- setNames(bike, c("date", "bike_count", "hour", "temp", "humidity", "wind_speed", "visibility", "dew_point_temp", "solar_radiation", "rainfall", "snowfall", "seasons", "holiday", "functioning_day"))

# Check the structure of the 'bike' dataset
str(bike)

# Summarize the total count of missing values in the 'bike' dataset
sum(is.na(bike))

# Display the dimensions (rows and columns) of the 'bike' dataset
dim(bike)
```

## Trasform Data
```{r}
# Convert specific columns to categorical variables for analysis
bike$hour <- factor(bike$hour)
bike$seasons <- factor(bike$seasons)
bike$holiday <- factor(bike$holiday)
bike$functioning_day <- factor(bike$functioning_day)

# Summarize the 'functioning_day' variable
summary(bike$functioning_day)
```




## Visualizations
### Bike Demand by Functioning Day
```{r}
# Create a box plot for Bike Demand by Functioning Day using ggplot2
fig1 <- ggplot(bike, aes(x = as.factor(functioning_day), y = bike_count)) +
  geom_boxplot(fill = "slateblue", alpha = 0.2) + # Customize boxplot appearance
  ggtitle("Fig 1. Bike Demand by Functioning Day") +  # Set the plot title
  xlab("Functioning Day") +  # Label for the x-axis
  ylab("Bike Count")  # Label for the y-axis

# Customize the plot title text appearance
fig1 + theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```
```{r}
# Deleting rows when it is non-functioning day
bike<-bike[!(bike$functioning_day=="No"),]

# removing unused columns
bike <- subset(bike, select = - c(functioning_day))
bike <- subset(bike, select = - c(date))

summary(bike)
```


### Hourly Bike Count Plot
```{r}
# Create a box plot for Hourly Bike Count Plot using ggplot2
fig2 <- ggplot(bike, aes(x = as.factor(hour), y = bike_count)) +
  geom_boxplot(fill = "blue", alpha = 0.2) +  # Customize boxplot appearance
  ggtitle("Fig 2. Hourly Bike Count Plot") +  # Set the plot title
  xlab("Hour") +  # Label for the x-axis
  ylab("Bike Count")  # Label for the y-axis

# Customize the plot title text appearance
fig2 + theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))

```

### Seasonal Bike Count Plot
```{r}
# Create a box plot for Seasonal Bike Count Plot using ggplot2
fig3 <- ggplot(bike, aes(x = as.factor(seasons), y = bike_count)) +
  geom_boxplot(fill = "blue", alpha = 0.2) +  # Customize boxplot appearance
  ggtitle("Fig 3. Seasonal Bike Count Plot") +  # Set the plot title
  xlab("Seasons") +  # Label for the x-axis
  ylab("Bike Count")  # Label for the y-axis

# Customize the plot title text appearance
fig3 + theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```

### Holiday Bike Count Plot
```{r}
# Create a box plot for Holiday Bike Count Plot using ggplot2
fig4 <- ggplot(bike, aes(x = as.factor(holiday), y = bike_count)) +
  geom_boxplot(fill = "blue", alpha = 0.2) +  # Customize boxplot appearance
  ggtitle("Fig 4. Holiday Bike Count Plot") +  # Set the plot title
  xlab("Holiday") +  # Label for the x-axis
  ylab("Bike Count")  # Label for the y-axis

# Customize the plot title text appearance
fig4 + theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))

```

### Histograms for numeric explanatory variables
```{r}
# Identify the numeric columns in the 'bike' data frame
continuous <- bike %>%
  select(where(is.numeric))

# Create histogram plots for numeric columns
histogram_plots <- lapply(names(continuous), function(column_name) {
  ggplot(data = bike, aes(x = .data[[column_name]])) +
    geom_histogram(fill = 'blue', color = 'black', bins = 30, alpha = 0.2) +
    labs(title = paste("Distribution of", column_name))
})

# Print or display the histogram plots
for (plot in histogram_plots) {
  print(plot)
}
```


### Histograms for numeric explanatory variables combined
```{r}
# Identify the numeric columns in the 'bike' data frame (excluding 'bike_count')
continuous <- bike %>%
  select(where(is.numeric)) %>%
  select(-bike_count)

# Create histogram plots for numeric columns
histogram_plots <- lapply(names(continuous), function(column_name) {
  ggplot(data = bike, aes(x = .data[[column_name]])) +
    geom_histogram(fill = 'blue', color = 'black', bins = 30, alpha = 0.2) +
    labs(title = paste("Distribution of", column_name))
})

# Arrange the plots in a 2x5 grid (2 rows and 5 columns)
grid.arrange(
  do.call(arrangeGrob, c(histogram_plots, ncol = 2)),
  ncol = 2
)
```

### Scatter Plot of Hour vs BikeCount for each Season
```{r}
ggplot(bike, aes(x = hour, y = bike_count, color = seasons)) +
  geom_point(size = 3) +
  labs(
    title = "Bike Count by Hour",
    x = "Hour",
    y = "Bike Count"
  )
```


# Method

## Part A


### Regression model with hour and seasons

```{r}
set.seed(32)
# Fit a linear regression model
bike.lm <- lm(bike_count ~ hour + seasons, data = bike)
```

### Assumptions
Let us check the normality assumption by investigating a normal Q-Q plot of the residuals.And homoscedasticity assumption by investigating fitted values of regression vs the residuals plot.

```{r}
# Set the plotting layout to one row and two columns
par(mfrow = c(1, 2))

# Plot 1: Normal QQ plot of Residuals
plot(bike.lm$fitted.values, bike.lm$residuals, main = "Normal QQ plot of Residuals")
abline(h = 0)  # Add a horizontal line at y = 0

# Plot 2: Residuals vs. Fitted Values
qqnorm(bike.lm$residuals, main = "Residuals vs Fitted Values", xlab = "Fitted Values", ylab = "Residuals")
qqline(bike.lm$residuals)  # Add a reference line to the QQ plot
```



### Log transformation of bike count
```{r}
# Fit a linear regression model with the log transformed bike_count
bike.lm.log <- lm(log(bike_count) ~ hour + seasons, data = bike)
```

## Assumptions for Log transformation model
Let us check the normality assumption by investigating a normal Q-Q plot of the residuals.And homoscedasticity assumption by investigating fitted values of regression vs the residuals plot.

```{r}
# Set the plotting layout to one row and two columns
par(mfrow = c(1, 2))

# Plot 1: Normal QQ plot of Residuals
plot(bike.lm.log$fitted.values, bike.lm.log$residuals, main = "Normal QQ plot of Residuals")
abline(h = 0)  # Add a horizontal line at y = 0

# Plot 2: Residuals vs. Fitted Values
qqnorm(bike.lm.log$residuals, main = "Residuals vs Fitted Values", xlab = "Fitted Values", ylab = "Residuals")
qqline(bike.lm.log$residuals)  # Add a reference line to the QQ plot
```



### Review the Model
```{r}
summary(bike.lm)
```


# Regression model for all variables

```{r}
bike.lm.all <- lm(bike_count ~., 
                  data = bike)
```

## Assumptions
Let us check the normality assumption by investigating a normal Q-Q plot of the residuals.And homoscedasticity assumption by investigating fitted values of regression vs the residuals plot.
```{r}
## look at the diagnostics
par(mfrow = c(1,2))
plot(bike.lm.all$fitted.values, bike.lm.all$residuals, main = "Normal QQ plot of Residuals")
abline(h=0)
qqnorm(bike.lm.all$residuals,main = "SResiduals vs Fitted Values", xlab= "Fitted Values", ylab= "Residuals")
qqline(bike.lm.all$residuals)
```

# Review the Model 3
```{r}
summary(bike.lm.all)
```


# Part B

## Principal Component Analysis 

### Covariance and Corelation
```{r}
# Identify the numeric columns in the 'bike' data frame
bike.continuous <- bike %>%
  select(temp, humidity, wind_speed, dew_point_temp, rainfall, snowfall, solar_radiation, visibility)

# Calculate covariances between columns in the 'stocks.subset' dataset, excluding the first column
bike.cov <- cov(bike.continuous)

# Display the covariance matrix
bike.cov

# Calculate the correlations between columns in the 'stocks.subset' dataset, excluding the first column
bike.cor <- cor(bike.continuous)

# Display the correlation matrix
bike.cor

```

### Perform Principal Component Analysis
```{r}
# Perform PCA on the 'stocks.subset' dataset
# The 'scale = T' argument scales the variables before performing PCA
colSums(is.na(bike.continuous))
bike.pca <- prcomp(~., data = bike.continuous, scale = T)

# Display the results of PCA
bike.pca

# Access the standard deviations of the principal components
bike.pca$sdev

# Access the rotation matrix, which contains the loadings of each variable on the principal components
bike.pca$rotation
```


### Scree Plot
```{r}
# Create a scree plot for the principal components
plot(bike.pca, type = 'l')

# Create a scree plot for the percentage of total variance explained by each principal component
plot(bike.pca$sdev^2/sum(bike.pca$sdev^2),
     ylim = c(0,1),
     xlab='Principal Component',
     ylab = 'Percentage of total variance',
     main = 'Scree Plot of Variance Accounted for each Principal Component',
     xaxt = "n", type="b")
axis(side=1, at=1:30)
```


### Calculate Percentage of Variance 
```{r}
# Calculate the percentage of variance explained by each principal component
percentage_of_variance <- (bike.pca$sdev^2) / sum(bike.pca$sdev^2)
percentage_of_variance

# Calculate the cumulative variance explained by the principal components
cumulative_var <- cumsum(bike.pca$sdev^2) / sum(bike.pca$sdev^2)
cumulative_var

# Create a plot of the cumulative variance explained by principal components
# The x-axis represents the number of components, and the y-axis represents the cumulative variance explained
plot(cumulative_var,  main = 'Cummulative Variance explained each Principal Component',
     xlab = "Principal Components", ylab = "Cumulative Variance Explained", type = "b")

```


### Biplot
```{r}
# Create a biplot for the PCA results
biplot(bike.pca)

# Create a scatterplot using ggbiplot
g = ggbiplot(bike.pca, obs.scale = 1, var.scale = 1, groups = bike.continuous$bike_count, ellipse = TRUE, circle = TRUE)

# Customize the scatterplot
g = g + scale_color_discrete(name = '')
g = g + theme(legend.direction = 'horizontal', legend.position = 'top')

# Print the scatterplot
print(g)
```


### The percentage contributions of the variables on PC1 
```{r}
# Extract the loadings for the first principal component (PC1)
loadings_PC1 <- bike.pca$rotation[, 1]

# Calculate the sum of absolute loadings for all variables
sum_all_loadings <- sum(abs(loadings_PC1))

# Calculate the percentage contributions of all variables to PC1
percentage_contributions <- round((abs(loadings_PC1) / sum_all_loadings) * 100, 2)

# Display the percentage contributions
percentage_contributions
```




### Cross Validation 
We will generate 10 candidate models selecting from seasons, hour and the first 3 variables that have a highest percentage contribution on PC1 (humidity, dew_point_temp, temp)

```{r}
# Define the function that relates X (where X is a design matrix) to the response Y.
model.fit <- function(x, y)
{
  return(lm(y ~ x - 1))
}

# Calculated the predicted/fitted values corresponding to this function.
predicted.values <- function(model.fit, x)
{
  return(x %*% model.fit$coefficients)
}

# Define the function that calculates prediction error for a given observation.  We will use squared residuals.
squared.residual <- function(Y, Y.hat)
{
  return((Y - Y.hat) ^ 2)
}


# 100 repetitions of cross-validation
ncrossval <- 100

# Construct a variable that includes the formulae corresponding to each of the candidate models.
formulae <- c("bike_count ~ seasons",
              "bike_count ~ hour",
              "bike_count ~ seasons + hour",
              "bike_count ~ seasons + hour + dew_point_temp + humidity + temp",
              "bike_count ~ dew_point_temp + humidity + temp",
              "bike_count ~ dew_point_temp + humidity",
              "bike_count ~ dew_point_temp + temp",
              "bike_count ~ humidity + temp",
              "bike_count ~ dew_point_temp",
              "bike_count ~ humidity",
              "bike_count ~ temp")


# Construct matrices for PRESS, MSE, and RMSE which have separate columns corresponding to the candidate models.  Each row stores values for a given repetition of cross-validation, so there are "ncrossval" rows.
PRESS.mat <- matrix(NA, nrow = ncrossval, ncol = length(formulae))
MSE.mat <- matrix(NA, nrow = ncrossval, ncol = length(formulae))
RMSE.mat <- matrix(NA, nrow = ncrossval, ncol = length(formulae))

# Re-set the random number generator seed.
set.seed(1234)

# Create an outer "for" loop that carries out "ncrossval" repetitions of cross-validation.
for(i in 1 : ncrossval)
{
  # Create an inner "for" loop that carries out cross-validation for each of the candidate models.
  for(j in 1 : length(formulae))
  {
    # Carry out ten-fold cross-validation for the given candidate model
    bike.cv <- crossval(model.matrix(as.formula(formulae[j]), data = bike), 
                        bike$bike_count, 
                        theta.fit = model.fit, 
                        theta.predict = predicted.values, 
                        ngroup = 10)  
    
    # Calculate PRESS, MSE, and RMSE for this candidate model and save to the appropriate cell in the matrix.
    PRESS.mat[i, j] <- sum((bike$bike_count - bike.cv$cv.fit) ^ 2)
    MSE.mat[i, j] <- PRESS.mat[i, j] / length(bike$bike_count)
    RMSE.mat[i, j] <- sqrt(MSE.mat[i, j])
  }
}

```

### Cross Validation Results
```{r}
# Print ten-fold cross-validation estimators of PRESS, MSE, and RMSE to the screen.
cv.results <- cbind(MODEL = formulae, PRESS=apply(PRESS.mat, MARGIN = 2, FUN = mean), 
                    MSE=apply(MSE.mat, MARGIN = 2, FUN = mean), RMSE=apply(RMSE.mat, MARGIN = 2, FUN = mean))
cv.results
```


### Model fit summary for the model with lowest MSE
```{r}
bike.cv.best <- lm(bike_count ~ seasons + hour + dew_point_temp + humidity + temp, data = bike)
summary(bike.cv.best)

# Calculate MSE and RMSE
MSE <- sum((bike$bike_count - bike.cv.best$fitted.values) ^ 2) / bike.cv.best$df.residual
RMSE <- sqrt(MSE)

# Print MSE and RMSE
c(MSE, RMSE)
```

